Formalizing Stack Safety as a Security Property

### 2022-09-01 submission5/onlyformalism.tex
Going to progressive introduction of concepts

>>>>

\section{The Basic Machine}

The building blocks of the machine are {\em words} and {\em registers}.
Words (\(\WORDS\)) are 64-bit integers ranged over by \(\word\) and, when used as addresses,
\(\addr\). The processor has 32 registers, ranged over in general by \(\reg \in \REGS\),
plus a program counter (\(\PCname\)). The following table divides
the registers into metavariables, some belonging to caller- or callee- sets:

Collectively addresses and registers are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \WORDS + \REGS\).
%
A {\em machine state} \(\mach \in \MACHS ::= \COMPONENTS \rightarrow \WORDS\)
is a mapping from state elements to values. For convenience, we abbreviate
multiple updates \(m[\component_0 \mapsto \word_0][\component_1 \mapsto \word_1]\dots\)
as \(m \llbracket A \mapsto B | C \rrbracket\), where \(A\) and \(B\)
are fomulae with free variables bound to sets in \(C\).
The base machine step function
\(\mach \stepstounder{} \mach' \in \MACHS \rightarrow \MACHS\) represents a single step.

The set of instructions, \(I \subseteq \WORDS\), corresponds to the RISC-V ISA.
We further introduce {\it macro-instructions}: sequences of
instructions that operate as if they were single, atomic pseudoinstructions.
A macro-instruction consists of both a list
of instructions and a pseudoinstruction that identifies its purpose in the
property. The pseudoinstructions, and the data that they carry, are as follows.

\judgment[Done]
         {\(\mach[\PCname] < a_{low} \lor \mach[\PCname] > a_{high}\)}
         {\(a_{low},a_{high} \vdash \mach \manystepstounder{} \mach\)}

\judgmenttwobrlong[Step]
                  {\(a_{low} \leq \mach[\PCname] < a_{high}\)}
                  {\(\mach \stepstounder{} \mach'\)}
                  {\(a_{low},a_{high} \vdash \mach' \manystepstounder{} \mach''\)}
                  {\(a_{low},a_{high} \vdash \mach \manystepstounder{} \mach''\)}

Now we're ready to introduce the overlay semantics.
Additionally, we keep track of security context information
specific to each stack over the course of the run.
This data is not relevant to execution,
it is only used to define the stack safety property. We define a set of
{\it security classes}:
\[sc \in SEC ::= \public | \sealed | \unsealed\]
A {\it security view} \(V \in \mathit{VIEW} ::= \COMPONENTS \rightarrow SEC\) divides the machine
elements in terms of their purpose relative to the active function.
The {\it initial view} \(V_0\) maps all stack locations and non-public registers
to \(\unsealed\) and all other locations and public registers to \(\public\).
A set of {\it targets} identify the addresses of the program counter or
stack pointer at a return point:
\[\begin{split}
tar \in TAR ::= & \mathit{Return} ~ a_{pc} ~ a_{sp} \\
& \mathit{Catch} ~ a_{pc} ~ a_{sp} \\
& \mathit{Done} \\
\end{split}\]
And a {\it stack context} is a non-empty stack of pairs of targets and views:
\[\context \in \CONTEXTS ::= \mathit{list} ~ (\mathit{TAR} \times \mathit{VIEW})\]

The initial stack context \(\context_0\) is \(\lbracket (\mathit{Done}, V_0) \rbracket\).

An {\it overlay state} \(\sigma\) consists of a pair of a machine state
and a stack contexts.
\[\sigma = (\mach, \context) \in \mathcal{S} ::= \MACHS \times \CONTEXTS\]

The labeled step function for overlay states
\(\sigma \stepstounder{\psi?} \sigma' \in
\mathcal{S} \rightarrow  (\Psi + \bot) \times \mathcal{S}\)
proceeds by an {\it overlay step}
if the program counter points to the start of a macro-instruction's instruction sequence.
Otherwise, it proceeds by the default step. The definitions of context steps
\(\constep{\psi} \in \CONTEXTS \times \Psi \rightarrow \CONTEXTS\)
will be defined for each pseudoinstruction as we introduce new topics.

\judgmentthreebrlong[Overlay]
                    {\((\bar{i}, \psi) \in J\)}
                    {\(\mach',\context \constep{\psi} \context'\)}
                    {\(\mach \manystepstounder{} \mach'\)}
                    {\(\mach[\PCname+n] = \bar{i}[n]\) for each \(n\) where \(0 \leq n < |\bar{i}|\)}
                    {\((\mach,\context)) \stepstounder{\psi} (\mach',\context')\)}

\judgment[Default]
         {\(\mach \stepstounder{} \mach'\)}
         {\((\mach,\context) \stepstounder{\bot} (\mach',\context)\)}

\paragraph*{Events and Traces}

Some macro-instructions may be interpreted as {\it events} that can be observed
from outside the system. For now we define events \(\obs \in \OBSS\) as calls, labeled
with the target address and the values of the arguments, or else as the silent
event \(\tau\):
\[\begin{split}
\obs \in \OBSS ::= & ~ \mathbf{callE} ~ a_{target} ~ \overline{w_{args}} \\
| & ~ \tau \\
\end{split}\]

\begin{figure*}
\[\mathit{observe} ~ \psi? ~ m =
\begin{cases}
  \mathbf{callE} ~ m[\PCname] ~ m \llbracket r | r \in \overline{r_{args}} \rrbracket
  & \textnormal{when } ~ \psi = \mathbf{call} ~ r_{target} ~ \overline{r_{args}} \\
  \mathbf{callE} ~ m[\PCname] ~ m \llbracket r | r \in \overline{r_{args}} \rrbracket
  & \textnormal{when } ~ \psi = \mathbf{tailcall} ~ r_{target} ~ \overline{r_{args}} \\
%  (\rho, \mathbf{returnE} ~ m[\PCname] & \textnormal{when } \psi = \mathbf{return} \\
  \tau & \textnormal{otherwise} \\
\end{cases}\]
\caption{Converting overlay steps to events}
\label{fig:observe}
\end{figure*}

\judgmenttwo{\((\mach,\context) \stepstounder{\psi?} (\mach',\context')\)}
            {\((\mach',\context') \hookrightarrow \obsT\)}
            {\((\mach,\context) \hookrightarrow \notfinished{\mathit{observe} ~ \psi? ~ \mach}{\obsT}\)}

We define another relation that takes a trace of a callee up to its return, or until
an exception is thrown, if either occurs.
We write this \(\context' \downarrow (\mach,\context) \hookrightarrow \obsT\), where \(\context\)
is the context of the caller.

\judgment{\(|\context| \leq |\context'|\)}
         {\(\context' \downarrow (\mach,\context) \hookrightarrow \tau\)}

\judgmenttwobrlong{\((\mach,\context) \stepstounder{\psi?} (\mach',\context')\)}
                  {\(|\context'| > |\context''|\)}
                  {\(\context'' \downarrow (\mach',\context') \hookrightarrow \obsT\)}
                  {\(\context'' \downarrow (\mach,\context) \hookrightarrow \notfinished{(\mathit{observe} ~ \psi? ~ \mach}{\obsT}\)}

We say that two event traces $\obsT_1$ and $\obsT_2$ are {\em similar},
written \(\obsT_1 \eqsim \obsT_2\), if the sequence of non-silent events
is the same. That is, we compare up to deletion of \(\tau\) events.

\begin{minipage}{.4\columnwidth}
  \judgment{}{\(\obsT \eqsim \obsT\)}
\end{minipage}
\begin{minipage}{.4\columnwidth}
  \judgment{\(\obsT_1 \eqsim \obsT_2\)}
           {\(\notfinished{\obs}{\obsT_1} \eqsim \notfinished{\obs}{\obsT_2}\)}
\end{minipage}

\begin{minipage}{.4\columnwidth}
  \judgment{\(\obsT_1 \eqsim \obsT_2\)}
           {\(\notfinished{\tau}{\obsT_1} \eqsim \obsT_2\)}
\end{minipage}
\begin{minipage}{.4\columnwidth}
  \judgment{\(\obsT_1 \eqsim \obsT_2\)}
           {\(\obsT_1 \eqsim \notfinished{\tau}{\obsT_2}\)}
\end{minipage}

%\begin{minipage}{.4\columnwidth}
%  \judgment{\(\obsT_1 \eqsim_\rho \obsT_2\)}
%           {\(\notfinished{(\rho',\obs)}{\obsT_1} \eqsim_\rho \obsT_2\)}
%\end{minipage}
%\begin{minipage}{.4\columnwidth}
%  \judgment{\(\obsT_1 \eqsim_\rho \obsT_2\)}
%           {\(\obsT_1 \eqsim_\rho \notfinished{(\rho',\obs)}{\obsT_2}\)}
%\end{minipage}

In this section we will present the context steps for calls and returns,
and show how they translate into properties. They are given formally in
Figure \ref{fig:callcontexts}.

For calls, we define a function, {\it update view}, that defines the view
of a caller in terms of that of the callee.
First, all stack locations below the stack pointer
are \(\unsealed\) -- fair game for the callee to allocate or write to, and
to read after writing. All \(\unsealed\) locations above the stack pointer
become \(\sealed\) -- their contents are both secret to and protected
from the callee. Argument and system registers are automatically marked
\(\mathit{public}\). All other locations above the stack pointer
retain their prior status (these will only be \(\sealed\) for now,
until we introduce shareable allocations shortly.)

To update the context for a call, we pair the updated view with the return target of the
call and push it to the stack. The return target will tell us whether a future
return is to the proper location with the stack pointer restored.
The view tells us how to treat the security of the callee's state, and as
it becomes more complicated, will benefit from the history that we maintain.
We take the head of a list, \(\mathit{hd}\), in the usual way, here a
total function because the context will never be empty.

On a return, we pop the topmost pair from the context that has
a \(\mathit{return}\), along with any other pairs we pass.

\begin{figure*}
\[\begin{split}
& \mathit{update\_view} ~ V ~ a_{sp} ~ \overline{r_{args}} \triangleq
\lambda \component .
\begin{cases}
  \public & \textnormal{else if } \component \in \REGS \textnormal{ and }
  \component \in \overline{r_{args}} \textnormal{ or } \component \in \mathit{PUBLIC} \\
  \unsealed & \textnormal{else if } \component \in \WORDS \textnormal{ and } \component \leq a_{sp} \\
  \sealed & \textnormal{else if } \component \in \WORDS \textnormal{ and } a_{sp} < \component \textnormal { and } V ~ \component = \unsealed \\
  V ~ \component & \textnormal{otherwise} \\
\end{cases} \\
\end{split}\]

\judgmenttwobrlong[CallC]
                  {\(\psi = \mathbf{call} ~ r_{target} ~ \overline{r_{args}}\)}
                  {\(\context' = (\mathit{Return} ~ (m[pc] + 4) ~ m[sp], V)::\context\)}
                  {\(V = \mathit{update\_view} ~ (snd ~ (hd ~ \context)) ~ m[sp] ~ \overline{r_{args}}\)}
                  {\(\mach, \context \constep{\psi} \context'\)}

\judgment[ReturnCRet]
         {\(\psi = \mathbf{return}\)}
         {\(\mach, (\mathit{return} ~ a_{pc} ~ a_{sp}, V)::\context \constep{\psi} \context\)}

\judgment[ReturnCDone]
         {\(\psi = \mathbf{return}\)}
         {\(m, [\mathit{Done},V] \constep{\psi} [\mathit{Done},V]\)}

\judgmenttwo[ReturnCSkip]
            {\(\psi = \mathbf{return}\)}
            {\(\mach, \context \constep{\psi} \context'\)}
            {\(\mach, (\_::\context) \constep{\psi} \context'\)}
